\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{authblk}
\usepackage{setspace}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}

\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}

\begin{document}
\title{\textbf{Indonesia Tariff Agreement After the US Supreme Court and President Trump Debacle: What It Means for Indonesia }}


\author[1]{Teguh Yudo Wicaksono \thanks{Faculty of Economics and Business, UIII}}
\author[1]{Rima P Artha}
\author[2]{Philips J. Vermonte\thanks{Faculty of Social Science, UIII}}
\affil[1]{Senior Fellow }
\affil[2]{Convenor}
\affil[1,2]{Indonesian Institute for Foreign Affairs (IIFA), Universitas Islam Internasional Indonesia (UIII)}


\date{February 22, 2026}
\onehalfspacing
\maketitle

%----------------------------------------------------------------------
\section*{ Summary}
%----------------------------------------------------------------------

\begin{itemize}
    \item On \textbf{February 19, 2026}, Indonesia and the US signed the Bilateral 
    Agreement of Reciprocal Trade (ART), reducing the reciprocal tariff from 32\% 
    to 19\% and granting zero-tariff exemptions to 1,819 tariff lines, including 
    palm oil, rubber, coffee, and cocoa.

    \item On \textbf{February 20, 2026}, the US Supreme Court ruled 6--3 in 
    \textit{Learning Resources, Inc.\ v.\ Trump} that IEEPA does not authorize 
    the President to impose tariffs---invalidating the entire legal basis of the ART.

    \item The Trump administration responded within hours by invoking Section 122 
    of the Trade Act of 1974 at 10\%, then escalating to \textbf{15\%---the 
    statutory maximum}---on February 21, 2026.

    \item Indonesia's \textbf{effective tariff rate (ETR) now stands at 20.3\%}, 
    exceeding the negotiated deal rate of 19.7\% by 0.6 percentage points.  
    \item \textbf{Hardest hit sectors---commodity exporters:}
    \begin{itemize}
        \item Palm oil (\$1.74B): tariff rises from 2\% (deal) to 17\%---an 
        increase of 15 percentage points, adding \$261 million in duties.
        \item Rubber (\$1.96B): tariff rises from 3\% to 18\%, adding 
        \$294 million in duties.
        \item Coffee (\$0.50B) and cocoa (\$0.32B): rise from 1.5--3\% to 
        16.5--18\%.
        \item Combined additional duty burden on commodities: \textbf{\$670 million} 
        relative to the deal.
    \end{itemize}

    \item \textbf{Modest relief---manufacturing exporters:}
    \begin{itemize}
        \item Textiles and apparel (\$4.22B, $\sim$4 million workers): tariff 
        declines from 29.1\% to 27\%.
        \item Footwear (\$2.57B): falls from 30\% to 26\%, saving \$103 million 
        annually.
        \item Furniture (\$1.57B): falls from 20.5\% to 16.5\%.
        \item Combined savings for manufacturing: approximately \textbf{\$500 million} 
        relative to the deal.
    \end{itemize}

    \item \textbf{Unaffected---steel and aluminum:} Section 232 tariffs remain 
    at 50\%, as they derive from separate legal authority and exceed the 15\% 
    Section 122 rate.

    \item Under the ART, Indonesia committed to: \$33 billion in US purchase 
    commitments (energy, Boeing aircraft, agricultural products); elimination 
    of tariffs on 99\% of US exports to Indonesia; and extension of 
    PT Freeport Indonesia's mining license from 2041 to 2061.

    \item These concessions were given in exchange for tariff terms that can 
    no longer be legally delivered.

   \end{itemize}

\newpage
%----------------------------------------------------------------------
\section{ Introduction}
%----------------------------------------------------------------------

The recent development on the tariffs imposed by President Trump on the U.S’ trading partners countries has raised controversies. Not only in the targeted countries, but the tariffs issues also generated controversies within the U.S. Domestic political battles between President Trump as the highest office holder in the executive branch of power and the Supreme Court of the United States (SCOTUS) in the judicial power continues and will certainly impact how various trade deal President Trump has signed with many countries, Indonesia included.

On April 2nd, 2025, by using his interpretation of the International Emergency Economic Powers (IEEPA), President Trump announced the so-called “Liberation Day” Executive Order that imposed 32\% reciprocal tariff on Indonesia. Not only Indonesia, but President Trump also imposed various rates to other countries he deemed as causing trade imbalance to the US. However, there was 90-day pause to give window for negotiation to which countries were responding. In late July 2025, Indonesian and the US negotiators reached a framework agreement in which the 32\% tariffs was then reduced to 19\%.


Eventually, Presidents Prabowo Subianto and Donald Trum signed the Bilateral Agreement of Reciprocal Trade (ART) on February 19th, 2026. It was met by the SCOTUS’ decision in the next day that invalidated President Trump’s use of IEEPA in the Liberation Day executive order.\footnote{President Trump also utilizes Section 232 Trade Expansion Act 1962 and Section 301 Trade Act 1974} In response, President Trump implemented the Section 122 Tariffs of 10 \% only to be raised within hours to 15\%.

The SCOTUS decision and President Trump’s response might be seen from two lenses. First, it shows the balancing act by a branch of power against another branch, particularly the powerful executive branch, as envisaged by the U.S Constitution. Since his first term, President Trump has shown a determination to fulfill his campaign promise of Make America Great Again (MAGA). In the context of the U.S foreign relations, MAGA is then translated into the “America First” principle. 

The America First principle has increased uncertainty in that it changes many multilateral foundations, both in economic and security spheres, that the U.S has as a matter of fact created in the past 80 years. Since President Trump took office, the U.S has acted more unilaterally. The U.S pushes many of her long-time allies, and punishes non-friendly countries, forcing them to accept the U.S’ side terms in many fronts. 

Second, Indonesia must watch very closely what is continuously developing in the U.S and beyond. While SCOTUS’ decision can be described as a democratic move to balance and restraint the U.S. President, for Indonesia the consequential tariff rate that ensue from this domestic “democratic balancing” dynamics can be economically hurting.

Countries like Indonesia who had signed bilateral tariff agreement with the U.S, before the SCOTUS’ decision and President Trump’s subsequent response, will now most likely have to recalculate the impacts of such a new decision on their economies. If, the countries found that the decision is hurting their economy more, they will seek to renegotiate to attain more favorable terms or at least maintain them in accordance to the already-signed agreement. 
Indonesia may need to do so as our preliminary calculations on the economic consequences of President Trump’s latest response (on February 21st, 2026) to SCOTUS’ decision (on February 20th, 2026). We found that Indonesian goods entering the U.S market will face higher tariffs compared to the tariffs deal that was signed by Presidents Prabowo and Trump on February 19th, 2026. 

In addition, the amount of time that is needed to renegotiate the terms brings uncertainty, which will also be detrimental to the Indonesian economy in general. Bearing in mind that President Trump, as the highest office holder of the U.S’ executive branch of power, still possess various policy tools at his disposal that he might use for his domestic political battles. 

President Trump’s subsequent policies in his domestic battles may, directly and indirectly, hurt our economy. Therefore, it must be noted that we are dealing with a moving target that may, or may not, changes quickly that presents a difficult and uncertain terrain to adapt and navigate. 

The next section of this short paper attempts to analyze the consequences of the fast-changing tariffs policy on Indonesia by calculating the possible effects on various sectors included in the tariff deal, considering the latest decision made by President Trump on February 21st, 2026, in his response to SCOTUS’ decision one day earlier. We also focus on the new effective tariff rate and which sectors that would be affected from the new tariffs direction. While the dynamics of tariffs policies still continues, this brief attempt to provide insights conditional on what has happened in the last few days.


%----------------------------------------------------------------------
\section{Context: The US--Indonesia Trade Relationship}
%----------------------------------------------------------------------


\subsection{Top Indonesian Export Sectors to the US}

Indonesia is America's 16th-largest goods trading partner. In 2024, US imports from Indonesia around \$28.05 billion. Meanwhile the US exports to Indonesia was around \$10.16 billion. For trade in goods, the US trade deficit with Indonesia is around \$17.9 billion (up 5.7\% from 2023) \cite{census2024,ustr2024}.

Indonesia holds dominant US market positions in several products: refined palm oil (85.1\%), industrial oleic acid (80.6\%), cinnamon (80.1\%), technically specified natural rubber (48\%), and crabs (46.8\%) \cite{comtrade}. Table \ref{tab:topexports} shows Indonesia's top exports based on the 2024 US import data \cite{usitc_dataweb}.

\begin{table}[h!]
\centering
\caption{Top Indonesian Export Sectors to the US (2024)}
\begin{tabular}{clccc}
\toprule
\textbf{Rank} & \textbf{Sector} & \textbf{HS Chapters} & \textbf{Value (\$B)} & \textbf{Share} \\
\midrule
1 & Electrical/Electronics & 85 & 4.51 & 16.1\% \\
2 & Footwear & 64 & 2.57 & 9.2\% \\
3 & Knitted Apparel & 61 & 2.20 & 7.8\% \\
4 & Woven Apparel & 62 & 2.02 & 7.2\% \\
5 & Rubber & 40 & 1.96 & 7.0\% \\
6 & Palm Oil/Fats & 15 & 1.74 & 6.2\% \\
7 & Furniture & 94 & 1.57 & 5.6\% \\
8 & Machinery & 84 & 1.01 & 3.6\% \\
9 & Mineral Fuels & 27 & 0.90 & 3.2\% \\
10 & Fish Preparations & 16 & 0.79 & 2.8\% \\
\bottomrule
\end{tabular}
\par\smallskip
\begin{minipage}{\linewidth}
\footnotesize
\textit{Note:} Values are customs value of U.S.\ general imports from Indonesia, calendar year 2024, aggregated at the HTS 2-digit chapter level. Share is each sector's value as a proportion of total U.S.\ imports from Indonesia (\$28.05 billion).\\
\textit{Source:} USITC DataWeb \cite{usitc_dataweb}; U.S.\ Census Bureau \cite{census2024}.
\end{minipage}
\label{tab:topexports}
\end{table}

\subsection{Chronology of the Tariff Crisis}

The Trump administration has changed their tariff policies across countries. Table \ref{tab:chronology} describes the chronology of the tariff crisis. It  describes the dynamics of the US tariffs and also the dynamics of the US-Indonesia trade relationship.

\begin{table}[htbp]
\centering
\caption{Chronology of the Tariff Crisis}
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Date} & \textbf{Event} \\
\midrule
April 2, 2025 & ``Liberation Day'' executive order: 32\% reciprocal tariff on Indonesia via IEEPA \\
April 9, 2025 & 90-day pause announced for countries entering negotiations \\
June 2025 & Section 232 tariffs on steel/aluminum doubled from 25\% to 50\% \cite{crs_232} \\
July 7, 2025 & Trump letter threatening reimposition of 32\% if no deal \\
July 22, 2025 & US--Indonesia framework agreement: reduced to 19\% \\
August 1, 2025 & 19\% reciprocal tariff takes effect \\
February 19, 2026 & Bilateral Agreement of Reciprocal Trade (ART) finalized \cite{art2026}; 1,819 tariff lines exempted; Freeport MoU signed \cite{freeport_mou} \\
February 20, 2026 & SCOTUS strikes down IEEPA tariffs 6--3 in \textit{Learning Resources v.\ Trump} \cite{scotus2026}; Trump invokes Section 122 at 10\% \\
February 21, 2026 & Trump raises Section 122 tariff from 10\% to 15\%---the statutory maximum \cite{s122_statute} \\
\bottomrule
\end{tabularx}
\par\smallskip
\begin{minipage}{\linewidth}
\footnotesize
\textit{Note:} Dates reflect the effective date of each executive action or announcement. All tariff rates are ad valorem.\\
\textit{Source:} Compiled from Executive Orders, White House fact sheets, \cite{scotus2026}, \cite{art2026}, \cite{freeport_mou}, \cite{crs_232}, and \cite{s122_statute}.
\end{minipage}
\label{tab:chronology}
\end{table}



\subsection{What Survives vs.\ What Falls}

Following the SCOTUS decision, all IEEPA reciprocal tariffs, including country specific bilateral deals under IEEPA were repealed. However, President Trump maintained section 232 tariffs, section 301 tariffs on China and also invoked section 122 with flat rate of 10\%. Table \ref{tab:legal survive} presents on regulations that survive and were repealed by the SCOTUS's decision.

\begin{table}[htbp]
\centering
\caption{Legal Authorities: What Survives vs.\ What Falls}
\begin{tabularx}{\textwidth}{XX}
\toprule
\textbf{Survives} & \textbf{Falls} \\
\midrule
Section 232 tariffs (steel, aluminum, autos) & All IEEPA reciprocal tariffs \\
Section 301 tariffs (China) & IEEPA universal baseline tariff \\
Anti-dumping / countervailing duties & Country-specific bilateral deals under IEEPA \\
Tariff Act of 1930 duties & \\
Section 122 (invoked at 10\%, raised to 15\%) & \\
\bottomrule
\end{tabularx}
\par\smallskip
\begin{minipage}{\linewidth}
\footnotesize
\textit{Note:} Classification reflects the immediate legal effect of the Supreme Court's 6--3 decision in \textit{Learning Resources, Inc.\ v.\ Trump} \cite{scotus2026} on February 20, 2026. ``Survives'' denotes authorities not derived from IEEPA and therefore unaffected by the ruling. Section 122 was invoked by executive order on the day of the ruling and raised to the 15\% statutory maximum on February 21, 2026.\\
\textit{Source:} Authors' analysis based on \cite{scotus2026}, \cite{s122_statute}, \cite{crs_232}, and \cite{crs_122}.
\end{minipage}
\label{tab:legal survive}
\end{table}

\newpage
%----------------------------------------------------------------------
\section{The Tariff Debacle and Its Implication to Indonesia}
%----------------------------------------------------------------------

 Following the SCOTUS ruling, President Trump utilized the Section 122 (\S122) and imposed global tariffs at 10\%. In just hours, the administration raised  tariff from 10\% to 15\%---the statutory maximum under 19 USC \S2132. This rapid development has brought ramifications for the global economy. While it may take time to fully assess the extent of its impacts on the global economy, the clear impact from the recent dynamics is on how to the US will implement tariffs for their trading counterparts.

 For Indonesia, what matter the most is on which tariff rates that the US will implement. In the following, we summarize the US tariff policies that are relevant to Indonesia. Specifically, table \ref{tab:etr} describes the evolution of tariff rates imposed by the US on Indonesia's exports. Following the Supreme Court decision, the initial Section 122 tariff of 10\% brought Indonesia's ETR down to 15.3\%. However, the subsequent escalation to 15\% on February 21 pushed the ETR to 20.3\%, which \textit{exceeds} the 19.7\% rate that Indonesia had negotiated under the bilateral deal. 

\begin{table}[htbp]
\centering
\caption{Effective Tariff Rates Across Scenarios}
\begin{tabular}{llcc}
\toprule
\textbf{Scenario} & \textbf{Period} & \textbf{ETR} & \textbf{Change from Baseline} \\
\midrule
1. Baseline & Pre-April 2025 & 5.0\% & -- \\
2. Liberation Day & April 2--9, 2025 & 36.6\% & +31.6 pp \\
3. Post-Deal & Aug 2025 -- Feb 19, 2026 & 19.7\% & +14.7 pp \\
4. Post-SCOTUS (10\% S122) & Feb 20, 2026 & 15.3\% & +10.3 pp \\
5. S122 at 15\% & Feb 21, 2026+ & 20.3\% & +15.3 pp \\
\bottomrule
\end{tabular}
\par\smallskip
\begin{minipage}{\linewidth}
\footnotesize
\textit{Note:} ETR is the trade-weighted average of applicable tariff rates across HTS 2-digit chapters, using 2024 U.S.\ import values as weights. Applicable rates incorporate MFN duties, Section 232 tariffs, and the relevant reciprocal or Section 122 rate under each scenario. The stacking rule applies: the higher of Section 232 or reciprocal/S122 governs each product; tariffs are not additive. Change from baseline is expressed in percentage points~(pp).\\
\textit{Source:} Authors' calculation based on USITC DataWeb \cite{usitc_dataweb}, USITC HTS \cite{usitc_hts}, and WTO Tariff Profiles \cite{wto_tariff}.
\end{minipage}
\label{tab:etr}
\end{table}

Perhaps, one that we do not expect is that the decision to raise the Section 122 tariff from 10\% to 15\% on February 21 has a counterintuitive consequence for Indonesia: a blunt, uniform tariff now imposes a \textit{higher} overall burden than the negotiated bilateral deal. Indonesia's total ETR rises from 15.3\% under the initial 10\% S122 to 20.3\% at the 15\% statutory maximum, surpassing the Post-Deal rate of 19.7\% by 0.6 percentage points. This outcome is driven by the structural difference between a flat tariff and a deal with exemptions. Under the bilateral agreement, palm oil (\$1.74B), rubber (\$1.96B), coffee (\$0.50B), and cocoa (\$0.32B) were fully exempted from the reciprocal tariff, paying only their low MFN rates of 1.5--3\%. Under Section 122, these same products face 15\% on top of MFN, representing a 12--15 percentage point increase from their deal-era rates.

The 5 percentage point escalation (from 10\% to 15\%) adds approximately \$1.40 billion in annual duties across all Indonesian exports. Every non-232 sector is affected uniformly: apparel goes from 22\% to 27\%, footwear from 21\% to 26\%, furniture from 11.5\% to 16.5\%, and commodity exports from 10--13\% to 15--18\%. Only products already covered by Section 232 (steel and aluminum at 50\%, autos at 25\%) are unaffected, since their 232 rates exceed the 15\% S122 rate and the stacking rule applies the higher of the two.

At 15\%, the administration has exhausted its unilateral authority under Section 122. The statute permits no further increase without congressional authorization. Combined with the 150-day time limit, this creates a peculiar policy landscape: the tariff is simultaneously at its maximum level and approaching its expiration date. For Indonesian exporters, this means the current 20.3\% ETR represents the worst-case scenario under existing executive authority---but it also means the clock is ticking. If Congress does not act to extend or replace Section 122 before the approximately July 20, 2026 expiration, non-232 tariffs would drop to zero, returning Indonesia's ETR close to the 5.0\% baseline.

\subsection{Winners and Losers: S122 at 15\% vs.\ the Deal}

Compared to the Post-Deal scenario, the current S122 at 15\% regime reshuffles the distribution of tariff burden across Indonesian export sectors. Manufacturing sectors that faced the full 19\% reciprocal tariff under the deal now face a lower 15\% S122 rate, while commodity sectors that were exempted under the deal now face the full 15\% with no relief.

While the overall ETR is around 20.3\%, it is important to note that not all sectors will face the same level of tariffs. Some sectors will be hit much harder due to higher tariff rates. In this part, we breakdown our tariff analysis into sectors. Table \ref{tab:sector impacts} presents the sector-level ETR comparison across all five scenarios. The ``S122 15\%'' column shows the current effective rate following the February 21 escalation. The ``vs.\ Deal'' column shows the change from the Post-Deal rate to the current S122 at 15\% rate.


\begin{table}[htbp]
\centering
\caption{Sector-Level ETR Comparison}
\small
\begin{tabular}{lcccccccr}
\toprule
\textbf{Sector} & \textbf{HS} & \textbf{(\$B)} & \textbf{Base} & \textbf{Lib.} & \textbf{Deal} & \textbf{S122} & \textbf{S122} & \textbf{vs.} \\
& & & & \textbf{Day} & & \textbf{10\%} & \textbf{15\%} & \textbf{Deal} \\
\midrule
Electronics & 85 & 4.51 & 1.8\% & 33.8\% & 14.1\% & 11.8\% & 16.8\% & +2.7 pp \\
Footwear & 64 & 2.57 & 11.0\% & 43.0\% & 30.0\% & 21.0\% & 26.0\% & $-$4.0 pp \\
Knit Apparel & 61 & 2.20 & 12.0\% & 44.0\% & 29.1\% & 22.0\% & 27.0\% & $-$2.1 pp \\
Woven Apparel & 62 & 2.02 & 12.0\% & 44.0\% & 29.1\% & 22.0\% & 27.0\% & $-$2.1 pp \\
\textbf{Rubber} & \textbf{40} & \textbf{1.96} & \textbf{3.0\%} & \textbf{35.0\%} & \textbf{3.0\%} & \textbf{13.0\%} & \textbf{18.0\%} & \textbf{+15.0 pp} \\
\textbf{Palm Oil} & \textbf{15} & \textbf{1.74} & \textbf{2.0\%} & \textbf{34.0\%} & \textbf{2.0\%} & \textbf{12.0\%} & \textbf{17.0\%} & \textbf{+15.0 pp} \\
Furniture & 94 & 1.57 & 1.5\% & 33.5\% & 20.5\% & 11.5\% & 16.5\% & $-$4.0 pp \\
Machinery & 84 & 1.01 & 1.5\% & 33.5\% & 20.5\% & 11.5\% & 16.5\% & $-$4.0 pp \\
\textbf{Coffee} & \textbf{09} & \textbf{0.50} & \textbf{1.5\%} & \textbf{33.5\%} & \textbf{1.5\%} & \textbf{11.5\%} & \textbf{16.5\%} & \textbf{+15.0 pp} \\
\textbf{Cocoa} & \textbf{18} & \textbf{0.32} & \textbf{3.0\%} & \textbf{35.0\%} & \textbf{3.0\%} & \textbf{13.0\%} & \textbf{18.0\%} & \textbf{+15.0 pp} \\
Steel & 72--73 & 0.30 & 26.5\% & 34.0\% & 52.0\% & 52.0\% & 52.0\% & 0 pp \\
Aluminum & 76 & 0.20 & 28.0\% & 35.0\% & 53.0\% & 53.0\% & 53.0\% & 0 pp \\
\bottomrule
\end{tabular}
\par\smallskip
\begin{minipage}{\linewidth}
\footnotesize
\textit{Note:} Bold rows indicate deal-exempted commodities that face the largest increases under S122 at 15\%. ETR for each sector is the trade-weighted average tariff at the HTS 2-digit chapter level. Applicable rates incorporate MFN duties plus the higher of Section 232 or reciprocal/S122 under each scenario (stacking rule). ``vs.\ Deal'' is the difference between S122 at 15\% and Post-Deal ETRs in percentage points~(pp). Import values are 2024 customs value.\\
\textit{Source:} Authors' calculation based on USITC DataWeb \cite{usitc_dataweb}, USITC HTS \cite{usitc_hts}, WTO Tariff Profiles \cite{wto_tariff}, and ART deal exemption schedule \cite{art2026}.
\end{minipage}
\label{tab:sector impacts}
\end{table}

Textiles and apparel (HS 61--62, \$4.22B) see their tariff drop from the deal-era effective rate of 29.1\% to 27\% under S122 at 15\%, a modest improvement of 2.1 percentage points. This sector employs approximately 4 million workers directly and supports 20 million livelihoods in Indonesia, so even small rate changes have significant employment implications. Footwear (HS 64, \$2.57B) benefits more substantially, with tariffs declining from 30\% to 26\%, saving an estimated \$103 million annually. Furniture (HS 94, \$1.57B) sees a 4 percentage point reduction from 20.5\% to 16.5\%.

On the losing side, the picture is stark. Rubber (HS 40, \$1.96B) faces tariffs of 18\% compared to just 3\% under the deal---a 15 percentage point increase that adds approximately \$294 million in annual duties. Palm oil (HS 15, \$1.74B) goes from 2\% to 17\%, adding \$261 million. Indonesia supplies 85\% of US refined palm oil, meaning buyers have limited substitution possibilities and the tariff burden will be partially absorbed by US consumers. Coffee and spices (HS 09, \$0.50B) rise from 1.5\% to 16.5\%, and cocoa (HS 18, \$0.32B) from 3\% to 18\%.

The net effect is an increase in the overall ETR of 0.6 percentage points compared to the deal (20.3\% vs.\ 19.7\%), translating to approximately \$170 million in additional total duties. This aggregate figure conceals a dramatic redistribution: manufacturing sectors save approximately \$500 million relative to the deal, while commodity sectors pay approximately \$670 million more.

The following are key takeaways on what we learned from the new tariff change.

\begin{itemize}
\item \textbf{At 15\% Section 122, Indonesia's ETR now exceeds the deal rate.} The escalation from 10\% to 15\% S122 raises Indonesia's overall ETR from 15.3\% to 20.3\%, surpassing the 19.7\% Post-Deal rate by 0.6 percentage points. A blunt, uniform tariff produces a higher burden than a targeted deal because deals include product exemptions.

\item \textbf{Every sector is worse off at 15\% S122 compared to 10\% S122.} Non-232 products that faced MFN~+~10\% now face MFN~+~15\%. Apparel goes from 22\% to 27\%, footwear from 21\% to 26\%, and palm oil from 12\% to 17\%.

\item \textbf{Key commodity exporters lose the most relative to the deal.} Palm oil, rubber, coffee, and cocoa exporters that enjoyed 0\% reciprocal tariff exemptions under the deal now face the uniform 15\% Section 122 tariff, representing a 12--15 percentage point increase.

\item \textbf{Section 232 tariffs on steel and aluminum remain unaffected} at 50\%, as they derive from a separate legal authority and exceed the 15\% S122 rate.

\item \textbf{Section 122 is at its statutory ceiling.} At 15\%, the administration has exhausted its authority under 19 USC \S2132. Any further increase requires congressional action.

\item \textbf{Section 122 is temporary.} The 150-day statutory limit expires around July 2026. Without congressional action, non-232 tariffs could revert toward baseline.

\item \textbf{The legal status of the Feb 19 deal is uncertain.} Its terms were negotiated under now-invalidated IEEPA authority, raising questions about enforceability of Indonesia's concessions.

\end{itemize}

\subsection{Indonesia's Trade Balance and its Implications}

The current 20.3\% ETR represents a 15.3 percentage point increase over the pre-crisis 5.0\% baseline, translating to approximately \$4.3 billion in additional duties beyond what Indonesian goods would face under normal MFN treatment. The distributional effects across sectors are uneven.

Indonesian textiles, footwear, and furniture face tariffs of 16.5--27\%, which is 5--16 percentage points above pre-crisis levels but still below the deal rates for these sectors. These labor-intensive industries, which together account for approximately \$10.4 billion (37\%) of exports to the US, remain under significant pressure but face a somewhat less punitive regime than under the IEEPA-era deal.

Palm oil and rubber, which together account for \$3.70 billion (13.2\% of exports to the US), are the hardest hit relative to the deal. They lose their zero-tariff exemptions and face the full 15\% S122 rate. For palm oil, where Indonesia dominates US supply (85.1\% market share), buyers have limited alternatives, potentially allowing partial pass-through of the tariff to US consumers. For rubber, with a 48\% US market share, the competitive impact is more severe as buyers may shift to synthetic substitutes or alternative natural rubber sources.

\subsection{Legal Status of the February 19 Deal}

The Agreement of Reciprocal Trade (ART) signed on February 19, 2026 \cite{art2026}, was negotiated under IEEPA authority that the Supreme Court invalidated the next day \cite{scotus2026}. This raises critical questions:

\begin{enumerate}
    \item \textbf{Is the deal enforceable?} The tariff reductions (19\% with 1,819 zero-tariff lines) were premised on IEEPA authority. Without that statutory basis, the US cannot legally implement the tariff structure as negotiated.

    \item \textbf{Are Indonesia's concessions still binding?} Indonesia agreed to:
    \begin{itemize}
        \item \$33 billion in US purchase commitments (energy, Boeing aircraft, agricultural products)
        \item Elimination of tariffs on 99\% of US products exported to Indonesia
        \item Extension of Freeport-McMoRan's mining license from 2041 to 2061
        \item Market access for US agriculture (soybeans, corn, wheat, cotton)
    \end{itemize}
    These concessions were given in exchange for tariff terms that can no longer be delivered. International trade law principles of reciprocity suggest Indonesia may have grounds to revisit or withdraw its commitments.

    \item \textbf{The Freeport MoU} \cite{freeport_mou}: Signed one day before the ruling, this agreement to extend PT Freeport Indonesia's mining license by 20 years represents a significant concession. Its legal status may depend on whether it was structured as a standalone agreement or as part of the broader trade package.
\end{enumerate}

\subsection{Section 122 Limitations}

Section 122's constraints create significant forward uncertainty:

\begin{itemize}
    \item \textbf{The 150-day clock}: Starting February 20, 2026, the tariff expires approximately July 20, 2026, unless Congress authorizes an extension. This creates a narrow window of tariff coverage.

    \item \textbf{No differentiation}: Unlike IEEPA tariffs, Section 122 cannot vary by country. Indonesia receives the same 15\% as Vietnam (which faced 46\% under IEEPA) or the EU (which faced 20\%). This eliminates the leverage for bilateral deal-making.

    \item \textbf{Statutory ceiling reached}: At 15\%, the rate is already at the statutory maximum. The administration has no remaining unilateral headroom to escalate further under this authority.
\end{itemize}


%----------------------------------------------------------------------
\section{Forward-Looking Scenarios}
%----------------------------------------------------------------------
It becomes clear to us that the internal dynamics in the US would have strong implications to the global economy, including Indonesia. Related to this context, the Indonesian government should monitor closely on what might potential happen in the next few months. Here we identify several possible scenarios and identify their implications to Indonesia.
\subsection{Scenario A: Congressional Action }

Congress could pass legislation granting the president new tariff authority, potentially through:
\begin{itemize}
    \item A revival of ``reciprocal trade'' authority under a new statutory framework
    \item Amendment to Section 301 to cover bilateral trade imbalances
    \item A standalone trade act with country-specific rate-setting power
\end{itemize}

\textbf{Implications for Indonesia}: If Congress acts before the 150-day Section 122 expiry, the administration could reimpose differentiated tariffs. The rate for Indonesia would depend on the new legislation's structure.

\subsection{Scenario B: Section 122 Expiry Without Replacement}

If Congress fails to act and the 150-day clock runs out (approximately July 20, 2026):
\begin{itemize}
    \item Non-232 products revert to MFN-only tariffs
    \item Indonesia's ETR drops back toward the 5.0\% baseline
    \item Steel/aluminum remain at 50\% (Section 232)
    \item The US loses most of its tariff leverage in bilateral negotiations
\end{itemize}

\subsection{Scenario C: Deal Restructuring Under Alternative Authority}

The administration could attempt to restructure the Indonesia deal under surviving authorities:
\begin{itemize}
    \item \textbf{Section 301}: Requires a formal investigation into unfair trade practices (time-consuming)
    \item \textbf{Section 201}: Requires US International Trade Commission injury finding (sector-specific)
    \item \textbf{Section 338}: Authorizes retaliatory tariffs up to 50\% for discriminatory treatment (rarely invoked)
\end{itemize}

\textbf{Implications}: Any restructuring would take months and likely result in different tariff structures. Indonesia would be negotiating from a stronger position, as the US lacks immediate leverage.

\subsection{Scenario D: Multiple Authority Layering}

Treasury Secretary Bessent indicated the administration would combine Section 232, Section 301, Section 122, Section 201, and Section 338 to maintain tariff pressure. This patchwork approach could:
\begin{itemize}
    \item Maintain elevated tariffs on specific sectors (steel/aluminum via 232, potentially textiles via 201/301)
    \item Create legal uncertainty and compliance complexity
    \item Face immediate legal challenges on the novel use of each authority
\end{itemize}

%----------------------------------------------------------------------
\section{Policy Recommendations}
%----------------------------------------------------------------------
Given the recent dynamics, the next question is what we should do and anticipate the rapid changing global trade amid higher uncertainties. In this part, we suggest several strategies for government and Indonesia businesses.

\subsection{For the Indonesian Government}

\begin{enumerate}
    \item \textbf{Re-strrategize and do not prematurely implement deal concessions.} With the legal basis for the ART invalidated, Indonesia should pause implementation of its commitments---particularly tariff reductions on US products and agricultural market access---until the deal's legal framework is clarified. Up until know, it is unclear on how the US tariffs will be implemented. In addition, we have limited knowledge on the implications of SCOTUS's ruling implications on the Indonesia-US trade deal.

    \item \textbf{Prepare for Section 122 expiry.} Develop contingency plans for three outcomes: congressional renewal, expiry without replacement, or replacement with alternative authority. Begin diplomatic engagement with US congressional leaders.

     \item \textbf{Diversify export markets.} While US market access remains important, the tariff instability reinforces the need to deepen trade relationships with RCEP partners, the EU (ongoing CEPA negotiations), and emerging markets.

    \item \textbf{Leverage the non-differentiation constraint.} Under Section 122, Indonesia faces the same 15\% as all other countries. While this is at the statutory maximum, it is still lower than what many competitors faced under IEEPA (e.g., Vietnam at 46\%, China at 34\%). Use this window to gain relative market share.
\end{enumerate}

\subsection{For Indonesian Businesses}

\begin{enumerate}
    \item \textbf{Textile and footwear exporters}: Take advantage of the reduced tariff (27\% vs.\ 29--30\% under the deal) to maintain competitive position. Accelerate shipments during the 150-day Section 122 window, as rates may change after expiry.

    \item \textbf{Palm oil and rubber exporters}: Prepare for the 15\% tariff by building inventory and exploring hedging strategies. Given Indonesia's dominant market position (85\% for palm oil, 48\% for rubber), some portion of the tariff can be passed through to US buyers. The 15 percentage point increase from deal-era rates is substantial, but the tariff is at its statutory maximum and temporary.

    \item \textbf{All exporters}: Engage US importers on duty drawback and refund possibilities for IEEPA-period tariffs. Consult customs brokers on filing CBP protests before statutory deadlines.

    \item \textbf{Manufacturers using US inputs}: Monitor whether Indonesia maintains its commitment to reduce tariffs on US products. If Indonesia pauses or reverses these concessions, input costs could remain lower.
\end{enumerate}

%----------------------------------------------------------------------
\section{Conclusions}
%----------------------------------------------------------------------

On February 20, 2026, the US Supreme Court struck down President Trump's use of the International Emergency Economic Powers Act (IEEPA) to impose reciprocal tariffs, just one day after Indonesia and the United States finalized a landmark bilateral trade agreement negotiated under that same authority. The following day, the administration raised the Section 122 tariff from 10\% to 15\%---the statutory maximum under 19 USC \S2132. This report documents recent development in the US. We also calculate the potential impacts of the current dynamics on Indonesian exports. One clear impact is on tariffs. Regarding the tariff, we calculate effective tariff rates (ETR) across five distinct scenarios to assess the shifting tariff landscape for Indonesian exports. We find that with the escalation from 10\% to 15\%, the effective tariff rate (ETR) for Indonesian export now reach 20.3\%, which is higher than what the US-Indonesia agreed on the bilateral negotion. The level of ETR is estimated by assuming that the new tariff direction is implemented on top of the current baseline. Without this assumption, the ETR for Indonesian exports is around 17.7\%. 

%----------------------------------------------------------------------
\section*{Appendix 1: Contexts}

%----------------------------------------------------------------------

\subsection*{The SCOTUS Decision: \textit{Learning Resources, Inc.\ v.\ Trump}}

In a 6--3 decision authored by Chief Justice Roberts, the Supreme Court held that ``IEEPA does not authorize the president to impose tariffs.'' The majority---Roberts, Sotomayor, Kagan, Gorsuch, Barrett, and Jackson---reasoned that:

\begin{itemize}
    \item IEEPA contains no reference to tariffs or duties
    \item Congress has never used the word ``regulate'' to authorize taxation
    \item Historically, there is no president that had previously interpreted IEEPA as conferring tariff power
    \item The Constitution grants tariff authority to ``Congress alone'' (Art.\ I, Sec.\ 8)
\end{itemize}

Thomas and Kavanaugh (joined by Alito) dissented.


\subsection*{The Section 122 Pivot and Escalation}

Within hours of the ruling, the Trump administration invoked Section 122 of the Trade Act of 1974 \cite{s122_statute,crs_122}, initially imposing a 10\% global tariff. The next day, February 21, the administration raised the rate to 15\%---the statutory maximum. Key constraints:

\begin{itemize}
    \item \textbf{Current rate:} 15\% ad valorem (the statutory ceiling)
    \item \textbf{Maximum duration:} 150 days without congressional authorization
    \item \textbf{No country differentiation:} Must apply uniformly
    \item \textbf{Trigger:} Presidential determination of ``large and serious'' balance-of-payments deficit
    \item \textbf{Historic first:} Section 122 has never before been used to impose tariffs
    \item \textbf{No further headroom:} At 15\%, the administration has exhausted its unilateral authority under this statute
\end{itemize}

%----------------------------------------------------------------------
\section*{Appendix 2: Calculating New Effective Tariff Rate for Indonesia}
%----------------------------------------------------------------------
In this part, we present our methods for recalculating the effective tariff rate for Indonesia's products exported to the US. 

\subsection*{Methodology}

Effective Tariff Rate (ETR) is calculated as:

\begin{equation}
\text{ETR} = \frac{\sum_{i} (\text{applicable\_rate}_i \times \text{import\_value}_i)}{\sum_{i} \text{import\_value}_i}
\end{equation}

\noindent where $i$ denotes each HTS-2 chapter, and the applicable rate for each chapter incorporates:

\begin{itemize}
    \item \textbf{MFN tariff rate} (baseline duty from the Harmonized Tariff Schedule \cite{usitc_hts,wto_tariff})
    \item \textbf{Section 232 tariff} (steel 50\%, aluminum 50\%, autos 25\%)
    \item \textbf{Reciprocal tariff} (32\% Liberation Day, 19\% Post-Deal) or \textbf{Section 122} (10\% initial, 15\% current)
\end{itemize}

\noindent\textbf{Stacking rule} (per Yale Budget Lab methodology \cite{yalebudgetlab_model}): Section 232 and reciprocal/Section 122 tariffs are mutually exclusive; the higher additional rate applies. Products are not double-charged.

\subsection*{Scenario Definitions}

We simulate changes in tariff rates across five scenarios that trace the evolution of US tariff policy toward Indonesia, from the pre-crisis baseline through the SCOTUS ruling and the Section 122 escalation to its statutory maximum.

\textbf{Scenario 1: Baseline (Pre-April 2025)}
\begin{itemize}
    \item MFN rates only for most products
    \item Section 232 at 25\% on steel (HS 72--73) and aluminum (HS 76)
    \item No reciprocal tariffs
\end{itemize}

\textbf{Scenario 2: Liberation Day (April 2--9, 2025)}
\begin{itemize}
    \item 32\% IEEPA reciprocal tariff on all Indonesian products
    \item Section 232 at 25\% (superseded by higher 32\% reciprocal)
    \item Effectively MFN + 32\% across the board
\end{itemize}

\textbf{Scenario 3: Post-Deal (August 2025 -- February 19, 2026)}
\begin{itemize}
    \item 19\% reciprocal tariff (reduced from 32\%)
    \item 1,819 tariff lines exempted at 0\% reciprocal (palm oil, rubber, coffee, cocoa, spices, semiconductor components)
    \item Textile TRQ: $\sim$10\% of HS 61--62 trade at MFN-only (conditional on US-sourced raw materials); remaining 90\% at MFN + 19\%
    \item Section 232 increased to 50\% on steel/aluminum (higher than 19\%, so 232 applies)
    \item Section 232 at 25\% on autos
\end{itemize}

\textbf{Scenario 4: Post-SCOTUS at 10\% S122 (February 20, 2026)}
\begin{itemize}
    \item 10\% Section 122 tariff (flat, global, no exemptions)
    \item Section 232 at 50\% on steel/aluminum (higher than 10\%, so 232 applies)
    \item Section 232 at 25\% on autos
    \item All IEEPA-based exemptions voided
\end{itemize}

\textbf{Scenario 5: S122 at 15\% --- Statutory Maximum (February 21, 2026+)}
\begin{itemize}
    \item 15\% Section 122 tariff (flat, global, no exemptions)---the statutory ceiling under 19 USC \S2132
    \item Section 232 at 50\% on steel/aluminum (higher than 15\%, so 232 applies)
    \item Section 232 at 25\% on autos (higher than 15\%, so 232 applies)
    \item No country differentiation; no product exemptions
\end{itemize}


%----------------------------------------------------------------------
\section*{Appendix 3: Data Sources and Validation}
\addcontentsline{toc}{section}{Appendix: Data Sources and Validation}
%----------------------------------------------------------------------

\subsection*{Cross-Validation}
\begin{itemize}
    \item \textbf{Yale Budget Lab}: ``State of U.S.\ Tariffs'' reports \cite{yalebudgetlab2026,yalebudgetlab2025jul,yalebudgetlab2025nov}
    \item \textbf{Yale baseline US ETR}: 2.4\% (January 2025)---Our Indonesia-specific baseline of 5.0\% is higher due to textiles/footwear concentration, consistent with expectations
    \item \textbf{Yale post-SCOTUS US ETR}: 9.1\% at 10\% S122---Our Indonesia-specific rate is 15.3\% at 10\% S122 (20.3\% at 15\% S122), reflecting S122 applied to Indonesia's high-MFN product mix
    \item \textbf{Yale Liberation Day increase}: +11.5 pp overall US---Consistent with the country-specific 32\% tariff driving Indonesia's ETR to 36.6\%
\end{itemize}

\subsection*{Reconciliation with Yale Budget Lab Model (Incremental ETR)}

In addition to this report's analysis, we ran the Yale Budget Lab's open-source Tariff-ETR model \cite{yalebudgetlab_model} directly on Indonesia's HS10-level import data. The Yale model produces different ETR figures because it measures a different quantity. Understanding the distinction is essential for interpreting the results.

\subsubsection*{What each calculation measures}

\begin{table}[htbp]
\centering
\caption{Methodological Differences: Total ETR vs.\ Incremental ETR}
\small
\begin{tabularx}{\textwidth}{lXX}
\toprule
& \textbf{This Report (Total ETR)} & \textbf{Yale Model (Incremental ETR)} \\
\midrule
Definition & Total tariff burden including MFN baseline duties & Additional tariff burden from executive actions only, excluding MFN \\
Includes MFN? & Yes & No \\
Granularity & HTS 2-digit ($\sim$27 chapter groups) & HTS 10-digit (4,921 product lines) \\
Deal exemptions & Indonesia-specific ART: 1,819 lines exempted (palm oil, rubber, coffee, cocoa, semiconductors); textile TRQ & General IEEPA config only; product exemptions apply to all countries (pharma, certain electronics) \\
Section 232 scope & 3 categories (steel, aluminum, autos) & 14 categories (steel, aluminum, autos, furniture, copper, softwood, others) \\
Import data & \$28.05B (rounded, multiple sources) & \$27.88B (Census Bureau API, consumption imports) \\
\bottomrule
\end{tabularx}
\par\smallskip
\begin{minipage}{\linewidth}
\footnotesize
\textit{Note:} The Yale model's incremental ETR measures only the tariff burden attributable to executive actions (Sections 232, IEEPA, and 122) above the pre-existing MFN schedule. This report's total ETR incorporates MFN baseline duties. The Yale model operates at HTS 10-digit granularity (4,921 tariff lines for Indonesia); this report aggregates at HTS 2-digit ($\sim$27 chapter groups). Section 232 scope differs because the Yale model classifies additional product categories (e.g., furniture, copper, softwood lumber) under Section 232 based on Federal Register definitions.\\
\textit{Source:} Authors' compilation based on Yale Budget Lab model \cite{yalebudgetlab_model} and this report's methodology (Section 4.1).
\end{minipage}
\end{table}

\subsubsection*{Side-by-side comparison}

\begin{table}[htbp]
\centering
\caption{ETR Reconciliation: Total vs.\ Incremental}
\begin{tabular}{lcccl}
\toprule
\textbf{Scenario} & \textbf{Total ETR} & \textbf{Incremental ETR} & \textbf{Diff.} & \textbf{Explanation} \\
\midrule
Post-Deal / 12-4 & 19.7\% & 18.0\% & $\sim$1.7 pp & MFN adds $\sim$5 pp; deal \\
& & & & exemptions subtract $\sim$3 pp \\
S122 at 10\% / 2-20 & 15.3\% & 12.7\% & $\sim$2.6 pp & Pure trade-weighted \\
& & & & average MFN rate \\
S122 at 15\% / 2-21 & 20.3\% & 17.7\% & $\sim$2.6 pp & Same MFN gap; no \\
& & & & exemptions in either calc \\
\bottomrule
\end{tabular}
\par\smallskip
\begin{minipage}{\linewidth}
\footnotesize
\textit{Note:} Total ETR is from this report's HTS 2-digit calculation including MFN baseline duties. Incremental ETR is from the Yale Budget Lab model \cite{yalebudgetlab_model} applied to Indonesia's HTS 10-digit import data, measuring only executive-action tariffs above MFN. The difference~(Diff.) captures the trade-weighted average MFN contribution, net of any divergence in deal-exemption treatment between the two models. Dates indicate the policy configuration used for each scenario.\\
\textit{Source:} Total ETR: authors' calculation; Incremental ETR: authors' calculation using the Yale Budget Lab model \cite{yalebudgetlab_model}.
\end{minipage}
\end{table}

\subsubsection*{Why the gaps differ across scenarios}

In the \textbf{S122 scenarios} (both 10\% and 15\%), there are no exemptions in either calculation. The consistent 2.6~pp difference reflects purely the trade-weighted average MFN rate for Indonesian products---this is the ``clean'' estimate of how much MFN contributes to the total ETR.

In the \textbf{Post-Deal} scenario, this report includes Indonesia-specific deal exemptions that reduce the total ETR: palm oil faces only 2\% (MFN) instead of 21\% (MFN~+ reciprocal), rubber faces 3\% instead of 22\%, and so on. The Yale model does not model these deal exemptions and therefore applies the full 19\% reciprocal rate to those products. This partially offsets the MFN inclusion, producing a smaller gap (1.7~pp vs.\ 2.6~pp).

\subsubsection*{Which measure to use}

\begin{itemize}
    \item \textbf{Total ETR} (this report) is appropriate for assessing the full tariff burden faced by Indonesian exporters, including standing MFN duties. It answers: ``What is the total duty rate on Indonesian goods entering the US?''
    \item \textbf{Incremental ETR} (Yale model) is appropriate for isolating the policy impact of executive tariff actions above the pre-existing MFN schedule. It answers: ``How much additional tariff burden do executive actions impose beyond the MFN baseline?''
\end{itemize}

\noindent Both measures are valid and complementary. For policy analysis comparing scenarios against each other, the incremental ETR provides a cleaner comparison of policy changes. For trade cost analysis and exporter decision-making, the total ETR reflects actual duty payments.





\begin{thebibliography}{99}

\bibitem{yalebudgetlab2026}
Yale Budget Lab, ``State of U.S.\ Tariffs,'' February 20, 2026.
\url{https://budgetlab.yale.edu/research/state-us-tariffs}

\bibitem{yalebudgetlab_model}
Yale Budget Lab, \textit{Tariff-ETRs}: Open-Source Effective Tariff Rate Model, GitHub.
\url{https://github.com/Budget-Lab-Yale/Tariff-ETRs}

\bibitem{yalebudgetlab2025jul}
Yale Budget Lab, ``State of U.S.\ Tariffs,'' July 23, 2025.

\bibitem{yalebudgetlab2025nov}
Yale Budget Lab, ``State of U.S.\ Tariffs,'' November 17, 2025.

\bibitem{census2024}
U.S.\ Census Bureau, \textit{Foreign Trade --- U.S.\ Trade in Goods with Indonesia}, 2024.
\url{https://www.census.gov/foreign-trade/balance/c5600.html}

\bibitem{ustr2024}
Office of the United States Trade Representative (USTR), Indonesia Country Commercial Guide / Trade Profile, 2024.

\bibitem{usitc_hts}
U.S.\ International Trade Commission (USITC), \textit{Harmonized Tariff Schedule of the United States}, 2025 revision.

\bibitem{wto_tariff}
World Trade Organization, \textit{World Tariff Profiles --- United States}, 2024/2025.

\bibitem{crs_232}
Congressional Research Service, ``Section 232 Investigations: Overview and Issues for Congress,'' IN12519.

\bibitem{crs_122}
Congressional Research Service, ``Section 122 of the Trade Act of 1974,'' R48435.

\bibitem{scotus2026}
\textit{Learning Resources, Inc.\ v.\ Trump}, 607 U.S.\ \_\_\_ (2026).

\bibitem{s122_statute}
Trade Act of 1974, \S 122, codified at 19 U.S.C.\ \S 2132.

\bibitem{art2026}
Agreement of Reciprocal Trade (ART), United States--Indonesia, signed February 19, 2026.

\bibitem{freeport_mou}
Memorandum of Understanding on the Extension of PT Freeport Indonesia's Mining License, signed February 19, 2026.

\bibitem{comtrade}
United Nations COMTRADE Database, via Trading Economics, U.S.--Indonesia bilateral trade data.

\bibitem{usitc_dataweb}
USITC DataWeb, HTS-level U.S.\ import statistics.
\url{https://dataweb.usitc.gov}

\end{thebibliography}

\end{document}
